---
title: "Assignment 1 - Language Development in ASD - Power and simulations"
author: "Cassandra Rempel"
date: "2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the third exciting part of the Language Development in ASD exercise

In this part of the assignment, we try to figure out how a new study should be planned (i.e. how many participants?) in order to have enough power to replicate the findings (ensuring our sample size is adequate, our alpha at 0.05 and our beta at 0.8):
1- if we trust the estimates of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
2- if we are skeptical of the current study. Report the power analysis and comment on what you can (or cannot) use its estimates for.
3- if we only have access to 30 participants. Identify the power for each relevant effect and discuss whether it's worth to run the study and why
The list above is also what you should discuss in your code-less report.


## Learning objectives

- Learn how to calculate statistical power
- Critically appraise how to apply frequentist statistical power

### Exercise 1

How much power does your study have (if your model estimates are quite right)?
- Load your dataset (both training and testing), fit your favorite model, assess power for your effects of interest (probably your interactions).
- Report the power analysis and comment on what you can (or cannot) use its estimates for.
- Test how many participants you would have to have to replicate the findings (assuming the findings are correct)

N.B. Remember that main effects are tricky once you have interactions in the model (same for 2-way interactions w 3-way interactions in the model). If you want to test the power of main effects, run a model excluding the interactions.
N.B. Check this paper: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12504
You will be using:
- powerSim() to calculate power
- powerCurve() to estimate the needed number of participants
- extend() to simulate more participants

```{r}
library(tidyverse, lme4)
library(simr)

#load the data
data <- read.csv("data_a2_p3.csv")
summary(data)

#Removing NAs from fixed effects because that makes sense
#called isolated to say isolated variables
isolated <- data %>% 
  dplyr::select(ID, CHI_MLU, VISIT, Diagnosis, Gender, Age) %>%  
  na.omit()

#Correcting classes, let's make them useable 
isolated$Diagnosis <- as.factor(isolated$Diagnosis)
isolated$Gender <- as.factor(isolated$Gender)
isolated$ID <- as.factor(isolated$ID)
str(isolated)
```


```{r}

#this section is used as the start to figure out power curves and how to utilize them effectively. This is the start point.
#model
m <- lmer(CHI_MLU ~ Gender + Age + Diagnosis*VISIT + I(VISIT^3) + (1 + VISIT | ID), isolated, REML = F)
summary(m)

#power simulation
power_VD <-  powerSim (m, simr::fixed("Diagnosis:VISIT"), nsim = 50)
power_VD

#PowerCurve for VISIT*Diagnosis - 50 simulation standard
powerCurveVD <- powerCurve(m, fixed("Diagnosis:VISIT"), along = "ID", nsim = 50)
powerCurveVD
plot(powerCurveVD)

```


### Exercise 2

How would you perform a more conservative power analysis?
- Identify and justify a minimum effect size for each of your relevant effects
- take the model from exercise 1 and replace the effects with the minimum effect size that you'd accept.
- assess the power curve by Child.ID, identifying an ideal number of participants to estimate each effect
- if your power estimates do not reach an acceptable threshold simulate additional participants and repeat the previous analysis
- Report the power analysis and comment on what you can (or cannot) use its estimates for.

```{r}
#same model as above
m2 <- lmer(CHI_MLU ~ Gender + Age + Diagnosis*VISIT + I(VISIT^3) + (1 + VISIT | ID), isolated, REML = F)

#Identify and justify a minimum effect size for each of your relevant effects
summary(data$CHI_MLU)

#CHI_MLU falls between 0 and 4.365, mean is 2.003 :

TD <- subset(data, Diagnosis =="TD")
ASD <- subset(data,Diagnosis =="ASD")

summary (TD$CHI_MLU) 
# Min. 1st Qu.  Median    Mean   3rd Qu.    Max.   
#0.5584  1.5706  2.3268  2.3082  2.8964  4.3648 
summary(ASD$CHI_MLU)
# Min. 1st Qu.  Median    Mean   3rd Qu.    Max. 
# 0.000   1.017   1.395   1.663   2.222   4.302

#how to calculate effect size, by taking the difference between the two groups (e.g., the mean of treatment group minus the mean of the control group) and dividing it by the standard deviation of one of the groups.

#1.663 - 2.3082 = -0.645 


#this is the fixed effect
fixef(m2)["DiagnosisTD:VISIT"] <- 0.1
minpowerCurveVD <- powerCurve(m2, fixed("Diagnosis:VISIT"), along = "ID", nsim = 50)
minpowerCurveVD

plot(minpowerCurveVD)

#produces too small a lower boundary, thus extension is necessary
```
#https://humburg.github.io/Power-Analysis/simr_power_analysis.html 

```{r}
#code for performing and ploting the extended curve, 100 participants
extendm2 <- extend(m2, along = "ID", n = 100)

extend_minpowerCurveVD <- powerCurve(extendm2, fixed("Diagnosis:VISIT"), along = "ID", nsim = 100)
extend_minpowerCurveVD

plot(extend_minpowerCurveVD)

```


### Exercise 3

Assume you have only the resources to collect 30 kids (15 with ASD and 15 TDs). Identify the power for each relevant effect and discuss whether it's worth to run the study and why

```{r}
#creating a small dataframe with 15 asd and 15 td for a total of 30
td <- isolated %>% 
  filter(Diagnosis == "TD")
asd <- isolated %>% 
  filter(Diagnosis == "ASD")
sampletd <- td$ID %>% 
  unique() %>% 
  sample(size = 15)
```


```{r}
#filter function exercise
sampleasd <- asd$ID %>% 
  unique() %>% 
  sample(size = 15)
smalltd <- isolated %>% 
  filter(ID %in% sampletd) 
smallasd <- isolated %>% 
  filter(ID %in% sampleasd)
small <- rbind(smalltd, smallasd)

#assumes the minimum effect, if not otherwise stated

#new model with the small df
m3 <- lmer(CHI_MLU ~ Gender + Age + Diagnosis*VISIT + I(VISIT^3) + (1 + VISIT | ID), small, REML = F)

#power calculation

sim_final <- powerSim (m3, simr::fixed("Diagnosis:VISIT"), nsim = 50)
sim_final
sim_powerCurve3 <- powerCurve(m3, fixed("Diagnosis:VISIT"), along = "ID", nsim = 50)
sim_powerCurve3

plot(sim_powerCurve3)

summary(m3)

#this is with the fixed effect

m4 <- lmer(CHI_MLU ~ Gender + Age + Diagnosis*VISIT + I(VISIT^3) + (1 + VISIT | ID), small, REML = F)
fixef(m4)["DiagnosisTD:VISIT"] <- 0.1
final_minCurve <- powerCurve(m4, fixed("Diagnosis:VISIT"), along = "ID", nsim = 50)
final_minCurve

plot(final_minCurve)
```
