---
title: "Assignment 2 - Language Development in ASD - Making predictions"
author: "Cassandra"
date: "sep 2020"
output:
  html_document: null
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the second exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time against some new data. Does the model generalize well?
Then we will learn to do better by cross-validating models and systematically compare them.

The questions to be answered (in a separate document) are:
1- Discuss the differences in performance of your model in training and testing data
2- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
3- Predict a new kid's performance (let's call him Bernie) and discuss it against expected performance of the two groups

## Learning objectives

- Critically appraise the predictive framework (contrasted to the explanatory framework)
- Learn the basics of machine learning workflows: training/testing, cross-validation, feature selections

## Let's go

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/5pc05mh5jwvdfjk/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/eegu8fea2entdqv/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/cf4p84mzn2p1bev/token_test.csv?dl=0

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())


```{r, include = FALSE}

#to start code, so we alwas access the file intended
#gives the path to the given file and the file
path_to_this_file <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(path_to_this_file)

#load the libraries
#the advantage of using pacman is it can install all packages and load them as well, if necessary
pacman::p_load(readr,dplyr,stringr,lmerTest,Metrics,caret, tidyverse, lme4, hydroGOF)
```


```{r}
# Load training Data
train_data1 <- read_csv("portfolio1_data.csv")
  
#- recreate the models you chose last time (just write the code again and apply it to Train Data). Both growth models and the best linear one
linear <- lmer(CHI_MLU ~ Gender + Age + VISIT*Diagnosis + (1 + VISIT | ID), train_data1, REML = F)
growth3 <- lmer(CHI_MLU ~ Gender + Age + VISIT*Diagnosis + I(VISIT^3) + (1 + VISIT | ID), train_data1, REML = F)

#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
trainlinear_rmse <- rmse(train_data1$CHI_MLU, predict(linear, train_data1, allow.new.levels = T)) 
#packages hydroGOF and Metrics are needed
traingrowth_rmse <- rmse(train_data1$CHI_MLU, predict(growth3, train_data1, allow.new.levels = T))
trainlinear_rmse
traingrowth_rmse

##- create the test dataset (apply the code from assignment 1 or my function to clean up the 3 test datasets)


# Test data
demo_test <- read_csv("demo_test.csv")
lu_test <- read_csv("LU_test.csv")
token_test <- read_csv("token_test.csv")

###LETS CLEAN UP THE DATA!!###

#renaming coloumns that are spelled differently, so they can be used
demo_test <- demo_test %>% 
  rename(ID = Child.ID,
         VISIT = Visit)
lu_test <- lu_test %>% 
  rename(ID = SUBJ)
token_test <- token_test %>% 
  rename(ID = SUBJ)

#making visit homogenous by extract the digit and making all the coloumns numerical
lu_test$VISIT <- str_extract(lu_test$VISIT, "\\d")
token_test$VISIT <- str_extract(token_test$VISIT, "\\d")

#removing problems in ID using a regular expression
lu_test$ID <- str_replace_all(lu_test$ID, "[[:punct:]]", "")
token_test$ID <- str_replace_all(token_test$ID, "[[:punct:]]", "")
demo_test$ID <- str_replace_all(demo_test$ID, "[[:punct:]]", "")

#choosing the variables we want
sub_demo_test <- select(demo_test, ID, VISIT, Diagnosis, Ethnicity, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization)
sub_token_test <- select(token_test, ID, VISIT, types_MOT, types_CHI, tokens_MOT, tokens_CHI)
sub_lu_test <- select(lu_test, ID, VISIT, MOT_MLU, CHI_MLU)

#merging two dataframes 
test_data <- merge(sub_demo_test,sub_lu_test)

#and merging with the last dataframe 
test_data <- merge(test_data, sub_token_test)

#creating a dataframe with the desired values from visit 1
visit1 <- subset(test_data, VISIT == 1, select = c(ID, ADOS, MullenRaw, ExpressiveLangRaw, Socialization))

#renaming the coloumns
visit1 <- visit1 %>% 
  rename(ADOS1 = ADOS) %>% 
  rename(MullenRaw1 = MullenRaw) %>% 
  rename(ExpressiveLangRaw1 = ExpressiveLangRaw) %>% 
  rename(Socialization1 = Socialization)

#merging the visit 1 values with the dataframe 
test_data <- merge(visit1, test_data)
#renaming coloumns
test_data <- test_data %>% 
  rename(verbalIQ = ExpressiveLangRaw,
         nonverbalIQ = MullenRaw,
         verbalIQ1 = ExpressiveLangRaw1,
         nonverbalIQ1 = MullenRaw1)
```


```{r}

#redoing values
test_data <- test_data %>% 
  mutate(Gender = ifelse(Gender == 1, "M", "F")) %>% 
  #determined by looking at the name Adam and his gender 
  mutate(Diagnosis = ifelse(Diagnosis == "A", "ASD", "TD"))

#saving the data
write.csv(test_data, file = "clean_test_data.csv")

#- test the performance of the models on the test data (Tips: google the functions "predict()")
testlinear_rmse <- rmse(test_data$CHI_MLU, predict(linear, test_data, allow.new.levels = T))
testgrowth_rmse <- rmse(test_data$CHI_MLU, predict(growth3, test_data, allow.new.levels = T))
testlinear_rmse
testgrowth_rmse

#- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

#Train data set is significantly better, therefore it is overfitted. The RMSE is signficantly lower, reflecting that it is overfitted. 

#OVerfitted is not good because it means the statistical methods are too specfically 

#Knows those select children very well but it isnt generalized enough because it only applies to the specific data set rather than a whole. 


```


[HERE GOES YOUR ANSWER]

### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model (Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)


- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data


```{r}
#removing X1 coloumn
train_data1 <- train_data1[,-1]
#renaming coloumns
train_data1 <- train_data1 %>% 
  rename(verbalIQ = ExpressiveLangRaw,
         nonverbalIQ = MullenRaw,
         verbalIQ1 = ExpressiveLangRaw1,
         nonverbalIQ1 = MullenRaw1)

#combining test and train data
TrainData <- rbind(train_data1, test_data)

#anonymizing
TrainData$ID <- as.factor(as.numeric(as.factor(TrainData$ID)))
```


```{r}
#Small world and large world.You will never have a large world if based on the small world data. Makes it important to cross validate and test the statistics

#THe more complex a model is, the easier it is to overfit. Greater details make it easier to make a data model that is too specific and not accurately testing the material.  

#Cross Validation. 

#- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

#choose some variations and see what model performs best, to evaluate language. Basically same model, different parameters and testing which makes the most sense. We're basically being testsed to understand the problems with complex models and the dangers of overfitting the data, which takes away from the results and their objective significants. 

#- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).
model_basic <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1 + VISIT | ID), TrainData, REML = F)

#redefining growth model for full dataset
growth_model <- lmer(CHI_MLU ~ Gender + Age + VISIT*Diagnosis + I(VISIT^3) + (1 + VISIT | ID), TrainData, REML = F)

#- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)
#basic model
k = 5
basic_trainRMSE <- rep(NA, k)
basic_testRMSE <- rep(NA, k)
folds = createFolds(unique(TrainData$ID), k = k, list = TRUE, returnTrain = FALSE)
i = 1

for (fold in folds) {
  train = subset (TrainData, !(ID %in% fold))
  test = subset (TrainData, ID %in% fold)
  model = model_basic
  test$prediction = predict(model, test, allow.new.levels = T)
  train$prediction = predict(model, train, allow.new.level = T) 
  basic_trainRMSE[i] = rmse(train$CHI_MLU, train$prediction)
  basic_testRMSE[i] = rmse(test$CHI_MLU, test$prediction)
  i = i + 1
}
basic_trainRMSE
basic_testRMSE

#growth model
growth_trainRMSE <- rep(NA, k)
growth_testRMSE <- rep(NA, k)
l = 1
for (fold in folds) {
  train = subset (TrainData, !(ID %in% fold))
  test = subset (TrainData, ID %in% fold)
  model = growth_model
  test$prediction = predict(model, test, allow.new.levels = T)
  train$prediction = predict(model, train, allow.new.level = T) 
  growth_trainRMSE[l] = rmse(train$CHI_MLU, train$prediction)
  growth_testRMSE[l] = rmse(test$CHI_MLU, test$prediction)
  l = l + 1
}
growth_trainRMSE
growth_testRMSE
#- Report the results and comment on them.
```


```{r}
## Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

#checking for collinearity
model_check1 <- data.frame(TrainData$nonverbalIQ, TrainData$verbalIQ)
print(correl_model_check1 <- round(cor(model_check1, use = "pair"), 2))
model_check2 <- data.frame(TrainData$types_CHI, TrainData$verbalIQ)
print(correl_model_check2 <- round(cor(model_check2, use = "pair"), 2))
model_check3 <- data.frame(TrainData$VISIT, TrainData$nonverbalIQ)
print(correl_model_check3 <- round(cor(model_check3, use = "pair"), 2))
model_check4 <- data.frame(TrainData$VISIT, TrainData$Age)
print(correl_model_check4 <- round(cor(model_check4, use = "pair"), 2))

#- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.
k=5
folds = createFolds(unique(TrainData$ID), k = k, list = TRUE, returnTrain = FALSE)


model1 <- lmer(CHI_MLU ~ Gender + Age + VISIT*Diagnosis + I(VISIT^2) + (1 + VISIT | ID), TrainData, REML = F)
P = 1
for (fold in folds) {
  train = subset (TrainData, !(ID %in% fold))
  test = subset (TrainData, ID %in% fold)
  model = model1
  test$prediction = predict(model, test, allow.new.levels = T)
  train$prediction = predict(model, train, allow.new.level = T) 
  growth_trainRMSE[P] = rmse(train$CHI_MLU, train$prediction)
  growth_testRMSE[P] = rmse(test$CHI_MLU, test$prediction)
  P = P + 1
}
growth_trainRMSE
growth_testRMSE
#working models in ^3
model2 <- lmer(CHI_MLU ~ Gender + Age + VISIT*Diagnosis + I(VISIT^3) + (1 + VISIT | ID), TrainData, REML = F)
G = 1
for (fold in folds) {
  train = subset (TrainData, !(ID %in% fold))
  test = subset (TrainData, ID %in% fold)
  model = model2
  test$prediction = predict(model, test, allow.new.levels = T)
  train$prediction = predict(model, train, allow.new.level = T) 
  growth_trainRMSE[G] = rmse(train$CHI_MLU, train$prediction)
  growth_testRMSE[G] = rmse(test$CHI_MLU, test$prediction)
  G = G + 1
}
growth_trainRMSE
growth_testRMSE
```


```{r}
#new model
model3 <- lmer(CHI_MLU ~ Gender + VISIT*Diagnosis + I(VISIT^3) + (1 + VISIT | ID), TrainData, REML = F) 
Y = 1
for (fold in folds) {
  train = subset (TrainData, !(ID %in% fold))
  test = subset (TrainData, ID %in% fold)
  model = model3
  test$prediction = predict(model, test, allow.new.levels = T)
  train$prediction = predict(model, train, allow.new.level = T) 
  growth_trainRMSE[Y] = rmse(train$CHI_MLU, train$prediction)
  growth_testRMSE[Y] = rmse(test$CHI_MLU, test$prediction)
  Y = Y + 1
} 
growth_trainRMSE
growth_testRMSE
AIC(model1, model2, model3)
```

model2 is best model 

[HERE GOES YOUR ANSWER]

### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)


```{r}
#isolate "Bernie" predict averages and expected developments. 

#how does it compare to the average TD and 

#is child below or above expectations
(test_data) 
#kids names are exposed, need to anonymize

#anonymizing
test_data$ID <- as.factor(as.numeric(as.factor(test_data$ID)))
#extracting child
TDkid <- filter(test_data, ID == 4) #random child
bernie <- filter(test_data, ID == 2) #actually bernie
#bernie's MLU compared to the average TD
td <-filter(TrainData, Diagnosis == "TD")
mean_td <- aggregate(td$CHI_MLU, list(td$VISIT), mean, na.rm = T)
difference <- bernie$CHI_MLU - mean_td$x
difference
#predicting bernie from chosen model (model 2, which is the same as our growth curve model)
bernie$predicted <- predict(model2, bernie, allow.new.levels = T)
bernie

#originally put "model" but we need to take from outside of the loop
#working with data inside of a loop and taking test data from outside of the loop
#cross validation loop only puts Bernie in context some of the time but not all of hte time, so we cannot build off of a loop

#why he is in the test set and not the training set
summary(bernie)
summary(TDkid)
bernie$CHI_MLU



#our prediction is undershooting, as he is outperforming our prediction
```

model is shit 
[HERE GOES YOUR ANSWER]

### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

